\textbf{Comparing and Analyzing the Effectiveness of Various String Matching Algorithms Across Different Domains}

### 2023-11-03 paper/main.tex
a misplaced underscore caused all this fuss

For instance, multiple studies have been recently published performing various comparative analyses of new string matching algorithms [4], and their applicability to a wide range of domains such as biological sequences [5], quantum computing [6], and instrusion detection systems [7]. These findings indicate the modern-day relevance and applicability of string matching algorithms and demonstrate its worthiness as a subject of investigation.

@@

\section{Theoretical Background} % Numbered section
\subsection{Topic Overview}
String matching algorithms (often used interchangeably with string or pattern searching
algorithms) are a subset of string algorithms. The term is composed of three parts:
“string”, “matching”, and “algorithm”. Strings refer to an abstract data type consisting
1
of a sequence of characters. Matching relates to it's function to identify specific patterns
within a larger dataset that match a criteria. As for algorithms, these can be defined as a set of instructions designed to carry out a function. These algorithms take in two
inputs: a pattern ($P$), which is a sequence of characters being searched for within a larger
sequence, a text ($T$). Within string matching problems, there exists two general variants
[2]:

\begin{enumerate}
	\item \textbf{Find occurences of a pre-defined pattern in a previously unseen dataset} \\
	Carried out using finite automata models [2] (an idealized machine used to recognize
	patterns in a given text; it accepts or rejects the input depending on whether the
	pre-defined pattern occurs in the text) or through the combinatorial properties of
	strings.
	\item \textbf{Find occurences of any identifiable patterns in a given text} \\
	Carried out using finite automata and binary trees.
\end{enumerate}

Linking into the mentioned problem variants, these algorithms can be further classified into two categories [1]: exact, and approximate string matching algorithms. As
its name suggests, the former refers to finding occurences of a pattern that match it to
the character, whereas the latter allows for some variation or deviation. For the purpose
of this investigation, the first problem variant, as well as an exact-string matching approach was chosen for all the algorithms, both for simplicity and control. Three exact
string matching algorithms were chosen for comparison based on their popularity: Naive,
Knuth-Morris-Pratt, and Rabin-Karp. Prior to their explanations, it is important to note
that the following notation will be used:

@@

[8] Also known as the “brute-force” approach, this algorithm compares the first indexes
of both $P$ and $T$. If $P_{i}$ equals $T_{j}$ then $i$ and $j$ are incremented and further compared. If
$P_{i}$ reaches $m - 1$ then a variable containing the number of occurences of $P$ is incremented
and $P_{i}$ is reset to $P_{i=0}$. On the other hand, if they do not match, then $P_{i}$ is reset and
$j$ incremented. This repeats until $T_{j=m-1}$ is reached. Since there is no pre-processing
phase, the only time complexity taken into account is during its matching phase, which
is expected to be $O(n \cdot m)$ [2]. The algorithm is better expressed in pseudocode:

@@

[2] [9] This algorithm differs from the Naive approach by implementing an LPS table
as part of the pre-processing phase. Its purpose is to keep track of the comparisons made;
after a mismatch, it is used to calculate where to begin the next match without needing
to reset $P$. The table is constructed by determining the largest prefix of $P$ that matches
its largest suffix -- the main idea is to identify how many characters can be skipped, by not
matching characters that have already been calculated to match (eliminating redundancy).
Although appearing complicated, this can be better explained through pseudocode and a
practical example represented through a truth table. Upon initalization, the LPS table is
an array of 0s of length $m$. For demonstration purposes, $P$ will equal "$ABABA$":

@@

[10] The most similar out of the two to the Naive approach, the Rabin-Karp algorithm
has a similar process of checking every substring for a match. Unlike the Naive approach
however, this algorithm compares “hash values”. These are calculated using rolling hash
functions, which allow for the removal and introduction of characters to the calculation,
eliminating the need to recalculate the value entirely from scratch. To perform the hashing
process on a substring, the following steps are carried out:

\begin{question}[\bf{Rolling Hash Function}]
\begin{enumerate}
	\item The base $(b)$ and modulus $(p)$ values are chosen (usually prime numbers).
	\item A hash value is initialised to 0 and the initial hash value of $P$ calculated (for each character in $P$, its contribution is added to the hash value as $c \cdot (b^{m-i-1}) \% p$.
	\item The hash value for the first substring of length $m$ in $T$ is calculated.
	\item Everytime $j$ is incremented, the contribution of the leftmost character is removed, the contribution of the rightmost character added.
	\item Compare the text and pattern matches each time; if they match then it is a \textbf{potential match} (rolling function outputs can be the same for different inputs). To verify the match, a character match is performed.

@@

\bibliographystyle{plain}
reference style
\bibliography{references}
\end{document}

>>>>

For instance, multiple studies have been recently published performing various comparative analyses of new string matching algorithms \parencite{4}, and their applicability to a wide range of domains such as biological sequences \parencite{5}, quantum computing \parencite{6}, and instrusion detection systems \parencite{7}. These findings indicate the modern-day relevance and applicability of string matching algorithms and demonstrate its worthiness as a subject of investigation.

@@

\section{Theoretical Background} % Numbered section
\subsection{Topic Overview}
\parencite{3} String matching algorithms (often used interchangeably with string or pattern searching
algorithms) are a subset of string algorithms. The term is composed of three parts:
“string”, “matching”, and “algorithm”. Strings refer to an abstract data type consisting
of a sequence of characters. Matching relates to it's function to identify specific patterns
within a larger dataset that match a criteria. As for algorithms, these can be defined as a set of instructions designed to carry out a function. These algorithms take in two
inputs: a pattern ($P$), which is a sequence of characters being searched for within a larger
sequence, a text ($T$). Within string matching problems, there exists two general variants
\parencite{2}:

\begin{enumerate}
	\item \textbf{Find occurences of a pre-defined pattern in a previously unseen dataset} \\
	Carried out using finite automata models \parencite{2} (an idealized machine used to recognize
	patterns in a given text; it accepts or rejects the input depending on whether the
	pre-defined pattern occurs in the text) or through the combinatorial properties of
	strings.
	\item \textbf{Find occurences of any identifiable patterns in a given text} \\
	Carried out using finite automata and binary trees.
\end{enumerate}

Linking into the mentioned problem variants, these algorithms can be further classified into two categories \parencite{1}: exact, and approximate string matching algorithms. As
its name suggests, the former refers to finding occurences of a pattern that match it to
the character, whereas the latter allows for some variation or deviation. For the purpose
of this investigation, the first problem variant, as well as an exact-string matching approach was chosen for all the algorithms, both for simplicity and control. Three exact
string matching algorithms were chosen for comparison based on their popularity: Naive,
Knuth-Morris-Pratt, and Rabin-Karp. Prior to their explanations, it is important to note
that the following notation will be used:

@@

\parencite{8} Also known as the “brute-force” approach, this algorithm compares the first indexes
of both $P$ and $T$. If $P_{i}$ equals $T_{j}$ then $i$ and $j$ are incremented and further compared. If
$P_{i}$ reaches $m - 1$ then a variable containing the number of occurences of $P$ is incremented
and $P_{i}$ is reset to $P_{i=0}$. On the other hand, if they do not match, then $P_{i}$ is reset and
$j$ incremented. This repeats until $T_{j=m-1}$ is reached. Since there is no pre-processing
phase, the only time complexity taken into account is during its matching phase, which
is expected to be $O(n \cdot m)$ \parencite{2}. The algorithm is better expressed in pseudocode:

@@

\parencite{2} \parencite{9} This algorithm differs from the Naive approach by implementing an LPS table
as part of the pre-processing phase. Its purpose is to keep track of the comparisons made;
after a mismatch, it is used to calculate where to begin the next match without needing
to reset $P$. The table is constructed by determining the largest prefix of $P$ that matches
its largest suffix -- the main idea is to identify how many characters can be skipped, by not
matching characters that have already been calculated to match (eliminating redundancy).
Although appearing complicated, this can be better explained through pseudocode and a
practical example represented through a truth table. Upon initalization, the LPS table is
an array of 0s of length $m$. For demonstration purposes, $P$ will equal "$ABABA$":

@@

\parencite{10} The most similar out of the two to the Naive approach, the Rabin-Karp algorithm
has a similar process of checking every substring for a match. Unlike the Naive approach
however, this algorithm compares “hash values”. These are calculated using rolling hash
functions, which allow for the removal and introduction of characters to the calculation,
eliminating the need to recalculate the value entirely from scratch. To perform the hashing
process on a substring, the following steps are carried out:

\begin{question}[\bf{Rolling Hash Function}]
\begin{enumerate}
	\item The base $(b)$ and modulus $(p)$ values are chosen (usually prime numbers).
	\item A hash value is initialised to 0 and the initial hash value of $P$ calculated (for each character in $P$, its contribution is added to the hash value as $c \cdot (b^{m-i-1}) \% p$).
	\item The hash value for the first substring of length $m$ in $T$ is calculated.
	\item Everytime $j$ is incremented, the contribution of the leftmost character is removed, the contribution of the rightmost character added.
	\item Compare the text and pattern matches each time; if they match then it is a \textbf{potential match} (rolling function outputs can be the same for different inputs). To verify the match, a character match is performed.

@@

\printbibliography
\end{document}
